---
title: "Comparing Search Algorithms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Search Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(modelminer)
```

## Overview

`modelminer` offers several search strategies for automated model building. They all use greedy forward selection as a backbone, but differ in how they construct and explore the candidate term pool:

| Method | Strategy |
|--------|----------|
| `greedy` | Single pre-built pool (all first-order, polynomial, and interaction terms); one pass |
| `greedy_alt` | Phased: (1) first-order, (2) polynomials on selected vars, (3) interactions among selected terms only |
| `greedy_alt_full` | Same phases as `greedy_alt`, but phase 3 uses ALL predictors for interactions |
| `greedy_alt_fb` | Same as `greedy_alt` with backward elimination steps within each phase |
| `greedy_alt_full_fb` | Same as `greedy_alt_full` with backward elimination steps within each phase |
| `forward_backward` | Single pre-built pool like `greedy`, but with backward elimination steps |

The phased methods (`greedy_alt*`) are more efficient but can miss signals that require interaction terms between variables with little individual main-effect signal. The `_full` variants address this by widening the interaction candidate pool in phase 3.

We illustrate these trade-offs with three simulation scenarios.

---

## Scenario 1: Main-Effects Signal

**Setup**: The response is a linear combination of two predictors; no interactions or polynomial terms matter.

```
y = 2*x1 - 1.5*x2 + noise
```

With only main effects driving the signal, phased methods should find the true model quickly. Standard `greedy` should find it too, but will evaluate more candidates along the way.

```{r scenario1-data}
set.seed(101)
n <- 80
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n); x4 <- rnorm(n)
x5 <- rnorm(n); x6 <- rnorm(n); x7 <- rnorm(n); x8 <- rnorm(n)
y  <- 2*x1 - 1.5*x2 + rnorm(n, sd = 1)
dat1 <- data.frame(y, x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario1-run, results="hide"}
configs1 <- list(
  greedy           = list(method = "greedy"),
  greedy_alt       = list(method = "greedy_alt"),
  greedy_alt_full  = list(method = "greedy_alt_full"),
  greedy_alt_fb    = list(method = "greedy_alt_fb"),
  greedy_alt_full_fb = list(method = "greedy_alt_full_fb"),
  forward_backward = list(method = "forward_backward")
)
cmp1 <- compare_methods(dat1, y, configs = configs1)
```

```{r scenario1-results}
cmp1$summary
```

**Interpretation**: All methods should identify `x1` and `x2` as the key predictors. The phased methods typically evaluate fewer total models because the candidate pool is pruned at each phase. `greedy` and `forward_backward` evaluate the full expanded pool from the start, which costs more comparisons but is not harmful here since the signal is clean.

---

## Scenario 2: Pure-Interaction Signal

**Setup**: The response depends only on the *product* of two predictors; neither variable has any meaningful marginal (main-effect) signal.

```
y = x1*x3 + noise
```

This is the critical test case distinguishing the phased methods. Because `x1` and `x3` have no individual signal:

- **`greedy_alt`**: Phase 1 selects predictors by main effects. `x1` and `x3` are not selected, so they never enter the interaction pool in phase 3. The interaction `x1:x3` is never evaluated — the true signal is **missed**.
- **`greedy_alt_full`**: Phase 3 uses ALL predictors for interactions, so `x1:x3` is always a candidate. The true signal is **found**.
- **`greedy`** and **`forward_backward`**: Build a single pool containing all interactions upfront, so `x1:x3` is evaluated from the start.

```{r scenario2-data}
set.seed(202)
n <- 80
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n); x4 <- rnorm(n)
x5 <- rnorm(n); x6 <- rnorm(n); x7 <- rnorm(n); x8 <- rnorm(n)
y  <- x1*x3 + rnorm(n, sd = 0.5)
dat2 <- data.frame(y, x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario2-run, results="hide"}
cmp2 <- compare_methods(dat2, y, configs = configs1)
```

```{r scenario2-results}
cmp2$summary
```

**Interpretation**: The `BestMetric` (AIC) for `greedy_alt` and `greedy_alt_fb` should be noticeably worse than the other methods — they settle for a near-null model because neither `x1` nor `x3` clears the bar individually. The `_full` variants and the single-pool methods recover the interaction and achieve a much lower AIC. This scenario exposes the fundamental limitation of the phased approach when main-effect signal is absent.

---

## Scenario 3: Mixed Signal

**Setup**: The response combines a main effect, a polynomial term, and an interaction between two predictors with only weak individual signal.

```
y = 2*x1 + x2^2 + x3*x4 + noise
```

This tests all three phases of the phased methods simultaneously. `x1` has a strong main-effect signal (phase 1 should pick it up); `x2^2` requires phase 2; and `x3*x4` requires phase 3. Whether `x3` and `x4` survive phase 1 determines whether `greedy_alt` can recover the interaction.

```{r scenario3-data}
set.seed(303)
n <- 80
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n); x4 <- rnorm(n)
x5 <- rnorm(n); x6 <- rnorm(n); x7 <- rnorm(n); x8 <- rnorm(n)
y  <- 2*x1 + x2^2 + x3*x4 + rnorm(n, sd = 1)
dat3 <- data.frame(y, x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario3-run, results="hide"}
cmp3 <- compare_methods(dat3, y, configs = configs1)
```

```{r scenario3-results}
cmp3$summary
```

**Interpretation**: Expect `greedy_alt_full` and the single-pool methods to fit the best models. `greedy_alt` may or may not recover `x3*x4` depending on whether `x3` and `x4` happen to clear phase 1 individually at this sample size. The `_fb` (forward-backward) variants sometimes improve on their non-backward counterparts by pruning spurious terms added early, leading to a cleaner final formula. Compare `ModelsEvaluated` across methods to appreciate the computational trade-off: more exhaustive search costs more evaluations but reduces the risk of missing important terms.

---

## Summary

The choice of method involves a **bias-variance trade-off in the search itself**:

- Use **`greedy_alt`** or **`greedy_alt_fb`** when you believe the true model's interactions involve predictors that already show marginal signal. These are the fastest and least likely to overfit by chance.
- Use **`greedy_alt_full`** or **`greedy_alt_full_fb`** when interactions among individually weak predictors are plausible. The wider phase-3 pool catches pure-interaction signals at modest additional cost.
- Use **`greedy`** or **`forward_backward`** when you want a single exhaustive pass over all candidate terms with no phasing assumptions. `forward_backward` adds backward pruning to remove terms that become redundant after later additions.

All methods return the same structure from `mine()` and are directly comparable via `compare_methods()`.
