---
title: "Comparing Search Algorithms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Search Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(modelminer)
```

## Methods

| Method | Interaction candidates | Notes |
|--------|----------------------|-------|
| `greedy` | `:` — interaction only | Standard greedy, single pool |
| `greedy_star` | `*` — interaction + main effects | Main effects always accompany their interaction |
| `greedy_alt` | `:` | Phased: first-order → polynomial → interactions among selected terms |
| `greedy_alt_full` | `:` | Same phases; phase 3 uses all predictors for interactions |
| `greedy_alt_fb` | `:` | `greedy_alt` with forward-backward within each phase |
| `greedy_alt_full_fb` | `:` | `greedy_alt_full` with forward-backward within each phase |
| `forward_backward` | `:` | Forward-backward over a single pool |

```{r configs}
configs <- list(
  greedy             = list(method = "greedy"),
  greedy_star        = list(method = "greedy_star"),
  greedy_alt         = list(method = "greedy_alt"),
  greedy_alt_full    = list(method = "greedy_alt_full"),
  greedy_alt_fb      = list(method = "greedy_alt_fb"),
  greedy_alt_full_fb = list(method = "greedy_alt_full_fb"),
  forward_backward   = list(method = "forward_backward")
)
```

---

## Scenario 1: Main-effects signal

```
y = 2*x1 - 1.5*x2 + noise
```

```{r scenario1-data}
set.seed(101)
n <- 80
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat1 <- data.frame(y = 2*x1 - 1.5*x2 + rnorm(n, sd = 1),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario1-run, results="hide"}
cmp1 <- compare_methods(dat1, y, configs = configs)
```

```{r scenario1-results}
cmp1$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 2: Pure-interaction signal

```
y = x1*x3 + noise   (no main effects)
```

```{r scenario2-data}
set.seed(202)
n <- 80
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat2 <- data.frame(y = x1*x3 + rnorm(n, sd = 0.5),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario2-run, results="hide"}
cmp2 <- compare_methods(dat2, y, configs = configs)
```

```{r scenario2-results}
cmp2$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 3: Mixed signal

```
y = 2*x1 + x2^2 + x3*x4 + noise
```

```{r scenario3-data}
set.seed(303)
n <- 80
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat3 <- data.frame(y = 2*x1 + x2^2 + x3*x4 + rnorm(n, sd = 1),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario3-run, results="hide"}
cmp3 <- compare_methods(dat3, y, configs = configs)
```

```{r scenario3-results}
cmp3$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```
