---
title: "Comparing Search Algorithms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Search Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(modelminer)

```

## Methods

| Method | Notes |
|--------|-------|
| `greedy` | Standard greedy, single pool |
| `greedy_star` | Like `greedy` but uses `*` for interactions, so main effects always accompany their interaction |
| `greedy_alt` | Phased: first-order → polynomial → interactions among selected terms |
| `greedy_alt_full` | Same phases; phase 3 uses all predictors for interactions |
| `greedy_alt_fb` | `greedy_alt` with forward-backward within each phase |
| `greedy_alt_full_fb` | `greedy_alt_full` with forward-backward within each phase |
| `forward_backward` | Forward-backward over a single pool |
| `backward` | Pure backward elimination from all predictors |
| `lasso` | L1 regularization via `cv.glmnet()`; returns models at `lambda.min` and `lambda.1se` |
| `lasso_path` | Full regularization path; records a model at each variable entry/exit point |

```{r configs}
configs <- list(
  greedy             = list(method = "greedy"),
  greedy_star        = list(method = "greedy_star"),
  greedy_alt         = list(method = "greedy_alt"),
  greedy_alt_full    = list(method = "greedy_alt_full"),
  greedy_alt_fb      = list(method = "greedy_alt_fb"),
  greedy_alt_full_fb = list(method = "greedy_alt_full_fb"),
  forward_backward   = list(method = "forward_backward"),
  backward           = list(method = "backward"),
  lasso              = list(method = "lasso"),
  lasso_path         = list(method = "lasso_path")
)
```

---

## Scenario 1: Main-effects signal

```
y = 2*x1 - 1.5*x2 + noise
```

```{r scenario1-data}
set.seed(101)
n <- 1000
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat1 <- data.frame(y = 2*x1 - 1.5*x2 + rnorm(n, sd = 1),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario1-run, results="hide"}
cmp1 <- compare_methods(dat1, y, configs = configs, verbose = FALSE)
```

```{r scenario1-results}
cmp1$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 2: Pure-interaction signal

```
y = x1*x3 + noise   (no main effects)
```

```{r scenario2-data}
set.seed(202)
n <- 1000
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat2 <- data.frame(y = x1*x3 + rnorm(n, sd = 0.5),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario2-run, results="hide"}
cmp2 <- compare_methods(dat2, y, configs = configs, verbose = FALSE)
```

```{r scenario2-results}
cmp2$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 3: Mixed signal

```
y = 2*x1 + x2^2 + x3*x4 + noise
```

```{r scenario3-data}
set.seed(303)
n <- 1000
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat3 <- data.frame(y = 2*x1 + x2^2 + x3*x4 + rnorm(n, sd = 1),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario3-run, results="hide"}
cmp3 <- compare_methods(dat3, y, configs = configs, verbose = FALSE)
```

```{r scenario3-results}
cmp3$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 4: Pure-quadratic signal

```
y = 3*x1^2 + noise   (no linear main effect)
```

```{r scenario4-data}
set.seed(404)
n <- 1000
x1<-rnorm(n); x2<-rnorm(n); x3<-rnorm(n); x4<-rnorm(n)
x5<-rnorm(n); x6<-rnorm(n); x7<-rnorm(n); x8<-rnorm(n)
dat4 <- data.frame(y = 3*x1^2 + rnorm(n, sd = 1),
                   x1, x2, x3, x4, x5, x6, x7, x8)
```

```{r scenario4-run, results="hide"}
cmp4 <- compare_methods(dat4, y, configs = configs, verbose = FALSE)
```

```{r scenario4-results}
cmp4$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## Scenario 5: Real data (mtcars)

```{r scenario5-run, results="hide"}
cmp5 <- compare_methods(mtcars, mpg, configs = configs, verbose = FALSE)
```

```{r scenario5-results}
cmp5$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

---

## CV metrics

The same scenarios repeated with `lm_loocv` (analytical leave-one-out CV via
the hat-matrix; no extra model fits, same speed as AIC) and `make_cv_metric`
(k-fold CV; refits the model k times per candidate, so too slow for n = 1000
and used here on mtcars only).

```{r cv-setup}
cv10 <- make_cv_metric(k = 10)
```

### Scenario 1

```{r cv-scenario1-run, results="hide"}
cmp1_loo <- compare_methods(dat1, y, configs = configs, metric = lm_loocv, verbose = FALSE)
```

```{r cv-scenario1-results}
cmp1_loo$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Scenario 2

```{r cv-scenario2-run, results="hide"}
cmp2_loo <- compare_methods(dat2, y, configs = configs, metric = lm_loocv, verbose = FALSE)
```

```{r cv-scenario2-results}
cmp2_loo$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Scenario 3

```{r cv-scenario3-run, results="hide"}
cmp3_loo <- compare_methods(dat3, y, configs = configs, metric = lm_loocv, verbose = FALSE)
```

```{r cv-scenario3-results}
cmp3_loo$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Scenario 4

```{r cv-scenario4-run, results="hide"}
cmp4_loo <- compare_methods(dat4, y, configs = configs, metric = lm_loocv, verbose = FALSE)
```

```{r cv-scenario4-results}
cmp4_loo$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Scenario 5 (mtcars) — lm_loocv

```{r cv-scenario5-loo-run, results="hide"}
cmp5_loo <- compare_methods(mtcars, mpg, configs = configs, metric = lm_loocv, verbose = FALSE)
```

```{r cv-scenario5-loo-results}
cmp5_loo$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Scenario 5 (mtcars) — 10-fold CV

```{r cv-scenario5-cv10-run, results="hide"}
cmp5_cv10 <- compare_methods(mtcars, mpg, configs = configs, metric = cv10, verbose = FALSE)
```

```{r cv-scenario5-cv10-results}
cmp5_cv10$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```
