---
title: "Getting Started with modelminer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with modelminer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`modelminer` automates stepwise model building -- iteratively adding or
removing polynomial expansions and interaction terms -- that most improve a model
selection metric. You supply a model function, a dataset, and a response
variable; `mine()` explores the candidate term space and returns the best
formula it finds, packaged as an S3 object with the standard
`print()`/`summary()`/`predict()`/`coef()` methods.

`mine()` is an S3 generic with a formula-first method (`mine(y ~ x, data)`)
and a data-first NSE method (`mine(data, y)`); both return the same kind
of object. The default metric is AIC and the default search is greedy
forward selection, but both are configurable. `mine()` accepts any model
function with a `formula` and `data` interface, which covers `lm`, `glm`,
and anything wrapped with `formula_wrap()`. See
`vignette("mine-fit-object")` for a tour of the returned object.

---

## Example 1: Linear model

The simplest case — finding predictors of fuel economy in `mtcars`:

```{r lm-example, results = "hide"}
library(modelminer)

lm_fit <- mine(mpg ~ ., data = mtcars,
               max_degree = 2, max_interact_vars = 2,
               verbose = FALSE)
```

```{r lm-result}
print(lm_fit)
```

The `$trace` data frame (also available as `$all_models` for backward
compatibility) records every formula evaluated during the search, including
the metric value at each step. This is useful for inspecting the search
path, diagnosing whether the algorithm stopped too early, or comparing how
metric values evolved:

```{r lm-trace}
head(lm_fit$trace[, "Formula", drop = FALSE])
nrow(lm_fit$trace)
```

---

## Example 2: Binomial GLM with an anonymous function

`mine()` is not limited to `lm`. Any model function with a `formula` and
`data` interface works. The `model_func` argument accepts anonymous functions,
which is the natural way to fix extra arguments (like `family`) while keeping
the signature compatible with what `mine()` expects.

Here we search for predictors of transmission type in `mtcars` (`am`: 0 =
automatic, 1 = manual):

```{r glm-example, results = "hide", warning = FALSE}
glm_fit <- mine(
  am ~ .,
  data              = mtcars,
  model_func        = \(formula, data) glm(formula, data, family = binomial),
  max_degree        = 1,
  max_interact_vars = 1,
  verbose           = FALSE
)
```

```{r glm-result}
formula(glm_fit)
```

Because `glm` returns an object that `AIC()` understands, no change to the
`metric` argument is needed. The anonymous function only needs to fix the
`family` argument that `mine()` has no way to know about.

> **Note on warnings:** Binomial GLMs commonly emit "fitted probabilities
> numerically 0 or 1 occurred" when a strong predictor (like `wt` in `mtcars`)
> separates the two classes nearly perfectly. This is a property of the data,
> not the search algorithm. The model and AIC values are still valid for
> selection purposes; it becomes a concern if you need reliable standard errors
> or p-values from the final model.

---

## Example 3: Lasso selection

For lasso/elastic-net variable selection, use `method = "lasso"` or
`method = "lasso_path"` instead of running stepwise search on top of
glmnet via `formula_wrap()`.  These native methods let regularization
handle selection in one shot, then refit with `lm` (or your `model_func`)
for the final model:

```{r lasso-example, eval = requireNamespace("glmnet", quietly = TRUE), results = "hide"}
# Requires the glmnet package.
# install.packages("glmnet")

lasso_fit <- mine(
  mpg ~ .,
  data              = mtcars,
  method            = "lasso",
  max_degree        = 2,
  max_interact_vars = 2,
  verbose           = FALSE
)
```

```{r lasso-result, eval = requireNamespace("glmnet", quietly = TRUE)}
formula(lasso_fit)
lasso_fit$best_metric
```

`method = "lasso_path"` walks the full regularization path and records
every model where the variable set changes, giving a richer view of the
selection landscape.  See `vignette("lasso-selection")` for details.

Extra arguments are forwarded to `cv.glmnet()` / `glmnet()`, so you can
set `alpha` (0 = ridge, 0.5 = elastic net, 1 = lasso), `nfolds`, `family`,
etc.

## Example 4: Matrix-based models with `formula_wrap()`

Some model functions don't accept a formula — `glmnet::cv.glmnet` takes a
predictor matrix `x` and response vector `y` instead. `formula_wrap()` adapts
such functions to `mine()`'s interface by extracting `x` and `y` from a data
frame using the formula.

> **Note:** For lasso variable selection, prefer `method = "lasso"` (Example
> 3) over wrapping `cv.glmnet` with `formula_wrap()`.  The wrapper runs
> stepwise search *on top of* lasso's own selection, which is redundant.
> `formula_wrap()` is still useful for other matrix-based models that don't
> have a native search method.

The anonymous function pattern from Example 2 applies here too: use it to
fix arguments the wrapper can't infer, such as the `family`.

```{r glmnet-example, eval = FALSE}
# Requires the glmnet package.
# install.packages("glmnet")

formula_cv_glmnet <- formula_wrap(
  \(x, y, ...) glmnet::cv.glmnet(x, y, ...)
)

# cv.glmnet needs at least one predictor column, so keep_all_vars = TRUE
# avoids passing an empty matrix for the intercept-only starting model.
glmnet_fit <- mine(
  mpg ~ .,
  data              = mtcars,
  model_func        = formula_cv_glmnet,
  metric            = \(m) min(m$cvm),   # minimum cross-validated MSE
  metric_comparison = min,
  keep_all_vars     = TRUE,
  max_degree        = 1,
  max_interact_vars = 1
)

formula(glmnet_fit)
```

`formula_wrap()` assumes the wrapped function uses `x` and `y` as argument
names (the `glmnet` convention). If your function uses different names, pass
them as `x_name` and `y_name`:

```{r formula-wrap-names, eval = FALSE}
# Example for a hypothetical function that uses "predictors" and "outcome"
formula_custom <- formula_wrap(my_model_func, x_name = "predictors", y_name = "outcome")
```

---

## Comparing configurations with `compare_methods()`

When experimenting with search settings — different algorithms, polynomial
degrees, or interaction depth — `compare_methods()` runs multiple
configurations and collects results in one call:

```{r compare-example, results = "hide"}
cmp <- compare_methods(
  mtcars, mpg,
  configs = list(
    greedy_linear   = list(method = "greedy",          max_degree = 1),
    greedy_poly     = list(method = "greedy",          max_degree = 3),
    forward_backward = list(method = "forward_backward", max_degree = 1)
  ),
  max_interact_vars = 1,
  verbose           = FALSE
)
```

```{r compare-result}
cmp$summary[, c("Config", "Formula", "BestMetric", "ModelsEvaluated")]
```

The `details` element of the result holds the full `mine()` output for each
configuration, so you can inspect `all_models` for any of them individually.
