---
title: "Getting Started with modelminer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with modelminer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`modelminer` automates forward model building by iteratively adding terms —
polynomial expansions and interaction terms — that most improve a model
selection metric. You supply a model function, a dataset, and a response
variable; `mine()` explores the candidate term space and returns the best
formula it finds.

The default metric is AIC and the default search is greedy forward selection,
but both are configurable. `mine()` accepts any model function with a
`formula` and `data` interface, which covers `lm`, `glm`, and anything
wrapped with `formula_wrap()`.

---

## Example 1: Linear model

The simplest case — finding predictors of fuel economy in `mtcars`:

```{r lm-example, results = "hide"}
# library(modelminer)
devtools::load_all()# modelminer isn't published at the moment, so load it from source.

lm_result <- mine(mtcars, mpg, max_degree = 2, max_interact_vars = 2)
```

```{r lm-result}
lm_result$Formula
```

The `all_models` data frame records every formula evaluated during the search,
including the metric value at each step. This is useful for inspecting the
search path, diagnosing whether the algorithm stopped too early, or comparing
how metric values evolved:

```{r lm-all-models}
head(lm_result$all_models[, "Formula", drop = FALSE])
nrow(lm_result$all_models)
```

---

## Example 2: Binomial GLM with an anonymous function

`mine()` is not limited to `lm`. Any model function with a `formula` and
`data` interface works. The `model_func` argument accepts anonymous functions,
which is the natural way to fix extra arguments (like `family`) while keeping
the signature compatible with what `mine()` expects.

Here we search for predictors of transmission type in `mtcars` (`am`: 0 =
automatic, 1 = manual):

```{r glm-example, results = "hide"}
glm_result <- mine(
  mtcars, am,
  model_func        = \(formula, data) glm(formula, data, family = binomial),
  max_degree        = 1,
  max_interact_vars = 1
)
```

```{r glm-result}
glm_result$Formula
# glm_result$all_models
```

Because `glm` returns an object that `AIC()` understands, no change to the
`metric` argument is needed. The anonymous function only needs to fix the
`family` argument that `mine()` has no way to know about.

---

## Example 3: Matrix-based models with `formula_wrap()`

Some model functions don't accept a formula — `glmnet::cv.glmnet` takes a
predictor matrix `x` and response vector `y` instead. `formula_wrap()` adapts
such functions to `mine()`'s interface by extracting `x` and `y` from a data
frame using the formula.

The anonymous function pattern from Example 2 applies here too: use it to
fix arguments the wrapper can't infer, such as the `family`.

```{r glmnet-example, eval = FALSE}
# Requires the glmnet package.
# install.packages("glmnet")

formula_cv_glmnet <- formula_wrap(
  \(x, y, ...) glmnet::cv.glmnet(x, y, ...)
)

# cv.glmnet needs at least one predictor column, so keep_all_vars = TRUE
# avoids passing an empty matrix for the intercept-only starting model.
glmnet_result <- mine(
  mtcars, mpg,
  model_func        = formula_cv_glmnet,
  metric            = \(m) min(m$cvm),   # minimum cross-validated MSE
  metric_comparison = min,
  keep_all_vars     = TRUE,
  max_degree        = 1,
  max_interact_vars = 1
)

glmnet_result$Formula
```

`formula_wrap()` assumes the wrapped function uses `x` and `y` as argument
names (the `glmnet` convention). If your function uses different names, pass
them as `x_name` and `y_name`:

```{r formula-wrap-names, eval = FALSE}
# Example for a hypothetical function that uses "predictors" and "outcome"
formula_custom <- formula_wrap(my_model_func, x_name = "predictors", y_name = "outcome")
```

---

## Comparing configurations with `compare_methods()`

When experimenting with search settings — different algorithms, polynomial
degrees, or interaction depth — `compare_methods()` runs multiple
configurations and collects results in one call:

```{r compare-example, results = "hide"}
cmp <- compare_methods(
  mtcars, mpg,
  configs = list(
    greedy_linear   = list(method = "greedy",          max_degree = 1),
    greedy_poly     = list(method = "greedy",          max_degree = 3),
    forward_backward = list(method = "forward_backward", max_degree = 1)
  ),
  max_interact_vars = 1
)
```

```{r compare-result}
cmp$summary[, c("Config", "Formula", "BestMetric", "ModelsEvaluated")]
```

The `details` element of the result holds the full `mine()` output for each
configuration, so you can inspect `all_models` for any of them individually.
