---
title: "Lasso and Regularization Path Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lasso and Regularization Path Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("glmnet", quietly = TRUE)
)
library(modelminer)
```

## Why lasso as a search method?

The stepwise methods in modelminer (greedy, forward-backward, etc.) add or
remove one term at a time, evaluating the metric at each step.  Lasso takes
a fundamentally different approach: it considers **all** candidate terms at
once and uses L1 regularization to shrink unimportant coefficients to zero.

modelminer provides two lasso-based methods:

| Method | What it does | Models returned |
|--------|-------------|-----------------|
| `"lasso"` | Runs `cv.glmnet()`, picks terms at `lambda.min` and `lambda.1se` | 2 (plus starting model) |
| `"lasso_path"` | Walks the full `glmnet()` regularization path, records every distinct variable set | One per variable entry/exit point |

Both methods build the full candidate pool (polynomials, interactions) from
`max_degree` and `max_interact_vars`, then let regularization decide which
terms to keep.  After lasso selects terms, `mine()` refits with the user's
`model_func` (default `lm`) and evaluates with the user's `metric` (default
`AIC`), so results are directly comparable with stepwise methods.

**Note:** `keep_all_vars` is ignored for both lasso methods.

---

## Basic usage

```{r basic, results = "hide"}
result <- mine(mtcars, mpg, method = "lasso",
               max_degree = 2, max_interact_vars = 2, verbose = FALSE)
```

```{r basic-result}
result$Formula
result$best_metric
```

The `all_models` table shows both the `lambda.min` and `lambda.1se`
selections:

```{r basic-all-models}
result$all_models
```

---

## Choosing lambda: `lambda.min` vs `lambda.1se`

By default, `method = "lasso"` returns the formula selected at `lambda.min`
(the lambda that minimizes cross-validated error).  The `lambda.1se` rule
picks the largest lambda within one standard error of the minimum --
producing a sparser, more conservative model.

Both models always appear in `all_models`.  The `lambda_rule` argument
controls which one becomes the returned `Formula`:

```{r lambda-rules, results = "hide"}
res_min <- mine(mtcars, mpg, method = "lasso", lambda_rule = "lambda.min",
                max_degree = 2, max_interact_vars = 1, verbose = FALSE)
res_1se <- mine(mtcars, mpg, method = "lasso", lambda_rule = "lambda.1se",
                max_degree = 2, max_interact_vars = 1, verbose = FALSE)
```

```{r lambda-rules-compare}
data.frame(
  rule    = c("lambda.min", "lambda.1se"),
  formula = c(deparse1(res_min$Formula), deparse1(res_1se$Formula)),
  AIC     = c(res_min$best_metric, res_1se$best_metric),
  n_terms = c(length(attr(terms(res_min$Formula), "term.labels")),
              length(attr(terms(res_1se$Formula), "term.labels")))
)
```

---

## The regularization path: `lasso_path`

`method = "lasso_path"` gives a richer picture.  Instead of returning just
two models, it walks the entire regularization path from most regularized
(empty model) to least regularized (all variables included) and records a
model at each point where the set of selected variables changes.

This is useful for understanding **how variables enter** -- which ones are
most important (enter first at high lambda) and which are marginal (enter
last at low lambda).

```{r path, results = "hide"}
path_result <- mine(mtcars, mpg, method = "lasso_path",
                    max_degree = 2, max_interact_vars = 2, verbose = FALSE)
```

```{r path-result}
path_result$all_models
```

```{r path-best}
path_result$Formula
path_result$best_metric
```

The `all_models` table reads like a story: the first row is the
intercept-only model, then each subsequent row adds (or occasionally drops)
a variable.  The best formula by AIC is typically somewhere in the middle --
not too sparse, not too dense.

---

## Comparing lasso with stepwise methods

```{r compare, results = "hide"}
cmp <- compare_methods(
  mtcars, mpg,
  configs = list(
    greedy           = list(method = "greedy"),
    forward_backward = list(method = "forward_backward"),
    backward         = list(method = "backward"),
    lasso            = list(method = "lasso"),
    lasso_path       = list(method = "lasso_path")
  ),
  max_degree = 2, max_interact_vars = 2, verbose = FALSE
)
```

```{r compare-result}
cmp$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

Lasso methods often find similar formulas to stepwise but evaluate far
fewer models -- regularization does the work in one shot rather than
iterating term-by-term.

---

## Elastic net and ridge regression

The `...` arguments in `mine()` are forwarded to `cv.glmnet()` (for
`"lasso"`) or `glmnet()` (for `"lasso_path"`).  This lets you control
the regularization mix via `alpha`:

| `alpha` | Regularization | Effect |
|---------|---------------|--------|
| 1 (default) | Pure lasso (L1) | Sparse: drives coefficients to exactly zero |
| 0.5 | Elastic net | Compromise: some sparsity, handles correlated predictors better |
| 0 | Ridge (L2) | No sparsity: keeps all variables, shrinks coefficients |

```{r elastic-net, results = "hide"}
# Elastic net (alpha = 0.5)
enet_result <- mine(mtcars, mpg, method = "lasso", alpha = 0.5,
                    max_degree = 1, max_interact_vars = 1, verbose = FALSE)
```

```{r elastic-net-result}
enet_result$Formula
```

Ridge regression (`alpha = 0`) never zeros out coefficients, so it selects
all variables.  This is useful as a baseline but not for variable selection.

Other `cv.glmnet` / `glmnet` arguments work too:

- `nfolds` -- number of CV folds (default 10)
- `family` -- response distribution (e.g., `"binomial"` for classification)

---

## Synthetic example: signal recovery

To see how lasso compares with stepwise on a problem with known structure:

```{r synthetic-data}
set.seed(42)
n <- 500
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n)
x4 <- rnorm(n); x5 <- rnorm(n); x6 <- rnorm(n)
# True model: y = 3*x1 - 2*x2 + x1*x3 + noise
y <- 3 * x1 - 2 * x2 + x1 * x3 + rnorm(n, sd = 1)
dat <- data.frame(y, x1, x2, x3, x4, x5, x6)
```

```{r synthetic-run, results = "hide"}
syn_cmp <- compare_methods(
  dat, y,
  configs = list(
    greedy     = list(method = "greedy"),
    fwd_bwd    = list(method = "forward_backward"),
    lasso      = list(method = "lasso"),
    lasso_path = list(method = "lasso_path")
  ),
  max_degree = 1, max_interact_vars = 2, verbose = FALSE
)
```

```{r synthetic-result}
syn_cmp$summary[, c("Config", "BestMetric", "ModelsEvaluated", "Formula")]
```

### Inspecting the lasso path

The `lasso_path` result shows how variables enter as regularization decreases:

```{r synthetic-path, results = "hide"}
syn_path <- mine(dat, y, method = "lasso_path",
                 max_degree = 1, max_interact_vars = 2, verbose = FALSE)
```

```{r synthetic-path-result}
syn_path$all_models
```

The order of variable entry often reveals the relative importance of
predictors: the strongest signals enter first (at high lambda), while
noise variables enter last (if at all).

---

## When to use each method

- **`"lasso"`** -- quick, principled variable selection. Good default when
  you want a single "best" model without iterating. Use `lambda_rule =
  "lambda.1se"` if you prefer parsimony.
- **`"lasso_path"`** -- when you want to see the full landscape of models
  from sparsest to densest, understand variable importance by entry order,
  or compare many nested models at once.
- **Stepwise methods** -- when you need fine-grained control over the search
  (e.g., forward-backward can correct early mistakes), or when using model
  functions that don't produce a numeric design matrix.

The lasso methods require the `glmnet` package:

```{r install, eval = FALSE}
install.packages("glmnet")
```
