---
title: "Working with mine() fit objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with mine() fit objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`mine()` is modelminer's single entry point for variable selection. It is an
S3 generic with two methods, so the same function name supports three
equivalent call forms:

```{r forms, eval = FALSE}
mine(mpg ~ ., mtcars)              # formula-first, the standard R contract
mine(mtcars, mpg)                   # data-first, NSE response
mtcars |> mine(mpg ~ wt + cyl)     # pipe-friendly, formula in slot 2
```

All three return the same classed object (`class(fit) == "mine"`), with the
same fields and the same S3 methods for `print()`, `summary()`, `coef()`,
`predict()`, `formula()`, and `plot()`.

---

## Install

```r
# From source (requires the remotes package)
remotes::install_git("https://github.com/jesseabrandt/modelminer.git")
```

---

## Fit a model

```{r fit}
library(modelminer)

fit <- mine(mpg ~ ., data = mtcars,
            method = "forward_backward",
            max_degree = 2, max_interact_vars = 2,
            verbose = FALSE)
```

`fit` is an S3 object of class `"mine"`:

```{r class}
class(fit)
```

## Print

`print()` shows the call, the search method, the selected formula, the final
metric, and the number of models the search evaluated. The full search trace
is *not* printed -- it can be hundreds of rows, so it lives in `fit$trace`
for inspection on demand.

```{r print}
print(fit)
```

## Summary

`summary()` returns a `summary.mine` object that combines the modelminer
metadata with the underlying model's summary. Print it to see both:

```{r summary}
summary(fit)
```

The summary object is a list, so individual pieces are accessible
programmatically:

```{r summary-parts}
s <- summary(fit)
s$best_metric
s$n_models
```

## Coefficients and predictions

`coef()` and `predict()` delegate to the fitted model:

```{r coef-predict}
coef(fit)

predict(fit, newdata = mtcars[1:5, ])
```

## Formula and trace

`formula()` returns the selected formula. The full search trace is in
`fit$trace`:

```{r formula-trace}
formula(fit)

head(fit$trace)
nrow(fit$trace)
```

## Extracting the underlying model

When you need to hand the fitted model to a function that expects a "real"
model object, use `extract_model()`:

```{r extract}
m <- extract_model(fit)
class(m)
```

## Restricting the search via the formula

A formula interface lets you scope the search to a subset of predictors
without first reshaping the data:

```{r restricted}
fit_small <- mine(mpg ~ wt + cyl + hp, data = mtcars,
                  method = "forward_backward",
                  max_degree = 2, max_interact_vars = 2,
                  verbose = FALSE)

formula(fit_small)
```

## The data-first shorthand

For interactive work where you already have the data in front of you, the
data-first form lets you name the response as a bare symbol -- every other
column in `data` becomes a candidate predictor:

```{r data-first}
fit_nse <- mine(mtcars, mpg,
                max_degree = 1, max_interact_vars = 1,
                verbose = FALSE)

formula(fit_nse)
```

This is equivalent to `mine(mpg ~ ., mtcars)` and returns the same kind of
object; use whichever reads more naturally at the call site.

## Legacy field names

If you have existing code that indexed the older `mine()` return value with
`$Formula` (capital F) or `$all_models`, those accessors still work -- the
classed `"mine"` object carries both the new and legacy field names:

```{r legacy}
identical(fit$formula,    fit$Formula)
identical(fit$trace,      fit$all_models)
```
