---
title: "Using modelminer with non-linear models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using modelminer with non-linear models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_all_pkgs <- all(vapply(
  c("tree", "rpart", "randomForest", "e1071"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_all_pkgs,
                      message = FALSE)
library(modelminer)
if (has_all_pkgs) {
  library(tree)
  library(rpart)
  library(randomForest)
  library(e1071)
}
```

## Does stepwise selection make sense for non-linear models?

Tree-based models and SVMs handle nonlinearity and variable selection
internally, so running a stepwise search on top of them is not obviously
useful. Whether it helps in practice likely depends on the data and the
model. This vignette doesn't try to settle that question — it just shows
that `mine()` works with these model functions and what the results look
like.

## Metrics

Each model family needs a metric that `mine()` can extract from a fitted
model.  Lower is better for all of these:

```{r metrics}
# tree: residual deviance (built-in generic)
# rpart: cross-validated relative error from the complexity table
rpart_xerror <- function(model) min(model$cptable[, "xerror"])
# randomForest: out-of-bag MSE (last tree in the sequence)
rf_oob_mse <- function(model) model$mse[model$ntree]
# svm: training residual MSE
svm_mse <- function(model) mean(model$residuals^2)
```

`tree` uses the standard `deviance()` generic, so no wrapper is needed.
`svm` stores residuals on the fitted object, so the wrapper just squares and
averages them.

## Compatibility notes

Not every formula feature works with every model function:

| Feature | `lm` | `tree` | `rpart` | `randomForest` | `svm` |
|---|---|---|---|---|---|
| Intercept-only (`y ~ 1`) | yes | yes | **no** | **no** | **no** |
| Polynomial (`I(x^2)`) | yes | yes | yes | only when `x` is a main effect | yes |
| Interaction (`x1:x2`) | yes | **no** | **no** | yes | yes |

`mine()` handles these differences automatically:

- When the intercept-only model fails (`rpart`, `randomForest`), `mine()`
  falls back to the best single-predictor model.
- Interaction terms that a model function rejects are skipped with a warning
  and the search continues.

For `rpart` and `tree`, set `max_interact_vars = 0` to avoid pointless
warnings from attempted interaction terms.

## Example: mtcars

### Decision tree (`tree`)

`tree()` accepts an intercept-only starting model, so `mine()` can begin
from `mpg ~ 1` and build up:

```{r tree-run, results = "hide"}
tree_result <- mine(
  mtcars, mpg,
  model_func        = tree,
  metric            = deviance,
  max_degree        = 2,
  max_interact_vars = 0
)
```

```{r tree-result}
formula(tree_result)
tree_result$best_metric
```

### Recursive partitioning (`rpart`)

`rpart()` cannot fit an intercept-only model, so `mine()` automatically
falls back to the best single-predictor starting point:

```{r rpart-run, results = "hide"}
rpart_result <- mine(
  mtcars, mpg,
  model_func        = rpart,
  metric            = rpart_xerror,
  max_degree        = 2,
  max_interact_vars = 0
)
```

```{r rpart-result}
formula(rpart_result)
rpart_result$best_metric
```

### Random forest

`randomForest()` also needs the single-predictor fallback.  Because the
forest is stochastic, we set a seed for reproducibility:

```{r rf-run, results = "hide"}
set.seed(42)
rf_result <- mine(
  mtcars, mpg,
  model_func        = randomForest,
  metric            = rf_oob_mse,
  max_degree        = 2,
  max_interact_vars = 2
)
```

```{r rf-result}
formula(rf_result)
rf_result$best_metric
```

### Support vector machine (`e1071::svm`)

`svm()` also cannot fit an intercept-only model, so the single-predictor
fallback kicks in.  SVM accepts both polynomial terms and interactions:

```{r svm-run, results = "hide"}
svm_result <- mine(
  mtcars, mpg,
  model_func        = svm,
  metric            = svm_mse,
  max_degree        = 2,
  max_interact_vars = 2
)
```

```{r svm-result}
formula(svm_result)
svm_result$best_metric
```

## Comparison: all five model families

Running the same search with `lm` (AIC) alongside the tree and SVM methods
shows how different model families select different features from the same data:

```{r lm-run, results = "hide"}
lm_result <- mine(
  mtcars, mpg,
  max_degree        = 2,
  max_interact_vars = 2
)
```

```{r comparison}
data.frame(
  model   = c("lm (AIC)", "tree (deviance)",
              "rpart (xerror)", "randomForest (OOB MSE)",
              "svm (MSE)"),
  formula = c(deparse1(formula(lm_result)),    deparse1(formula(tree_result)),
              deparse1(formula(rpart_result)),  deparse1(formula(rf_result)),
              deparse1(formula(svm_result))),
  n_terms = c(length(attr(terms(formula(lm_result)), "term.labels")),
              length(attr(terms(formula(tree_result)), "term.labels")),
              length(attr(terms(formula(rpart_result)), "term.labels")),
              length(attr(terms(formula(rf_result)), "term.labels")),
              length(attr(terms(formula(svm_result)), "term.labels")))
)
```

The model families pick noticeably different subsets. `lm` pulls in
interactions and polynomials; the single-tree methods stay compact;
`randomForest` and `svm` fall somewhere in between.

## Backward elimination

The examples above all use greedy forward selection — starting from a small
model and adding terms.  `method = "backward"` does the opposite: it starts
from all predictors and removes the least useful one each round until no
removal improves the metric.

Backward elimination is a natural fit for tree-based models because it
sidesteps the intercept-only fallback entirely — the search always begins
with every predictor in the formula.

```{r backward-lm, results = "hide"}
bwd_lm <- mine(
  mtcars, mpg,
  method            = "backward",
  max_degree        = 1,
  max_interact_vars = 0
)
```

```{r backward-tree, results = "hide"}
bwd_tree <- mine(
  mtcars, mpg,
  model_func        = tree,
  metric            = deviance,
  method            = "backward",
  max_degree        = 1,
  max_interact_vars = 0
)
```

```{r backward-rpart, results = "hide"}
bwd_rpart <- mine(
  mtcars, mpg,
  model_func        = rpart,
  metric            = rpart_xerror,
  method            = "backward",
  max_degree        = 1,
  max_interact_vars = 0
)
```

```{r backward-rf, results = "hide"}
set.seed(42)
bwd_rf <- mine(
  mtcars, mpg,
  model_func        = randomForest,
  metric            = rf_oob_mse,
  method            = "backward",
  max_degree        = 1,
  max_interact_vars = 0
)
```

```{r backward-svm, results = "hide"}
bwd_svm <- mine(
  mtcars, mpg,
  model_func        = svm,
  metric            = svm_mse,
  method            = "backward",
  max_degree        = 1,
  max_interact_vars = 0
)
```

### Forward vs backward on mtcars

To compare fairly, we run forward selection with matching settings
(`max_degree = 1`, `max_interact_vars = 0`):

```{r fwd-for-comparison, results = "hide"}
set.seed(42)
fwd_lm <- mine(mtcars, mpg, max_degree = 1, max_interact_vars = 0)
fwd_tree <- mine(mtcars, mpg, model_func = tree, metric = deviance,
                 max_degree = 1, max_interact_vars = 0)
fwd_rpart <- mine(mtcars, mpg, model_func = rpart, metric = rpart_xerror,
                  max_degree = 1, max_interact_vars = 0)
fwd_rf <- mine(mtcars, mpg, model_func = randomForest, metric = rf_oob_mse,
               max_degree = 1, max_interact_vars = 0)
fwd_svm <- mine(mtcars, mpg, model_func = svm, metric = svm_mse,
                max_degree = 1, max_interact_vars = 0)
```

```{r forward-vs-backward}
full_mtcars <- mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb

set.seed(42)
data.frame(
  model   = c("lm (AIC)", "tree (deviance)",
              "rpart (xerror)", "randomForest (OOB MSE)",
              "svm (MSE)"),
  forward_formula  = c(deparse1(formula(fwd_lm)), deparse1(formula(fwd_tree)),
                       deparse1(formula(fwd_rpart)), deparse1(formula(fwd_rf)),
                       deparse1(formula(fwd_svm))),
  backward_formula = c(deparse1(formula(bwd_lm)), deparse1(formula(bwd_tree)),
                       deparse1(formula(bwd_rpart)), deparse1(formula(bwd_rf)),
                       deparse1(formula(bwd_svm))),
  fwd_metric  = c(fwd_lm$best_metric, fwd_tree$best_metric,
                  fwd_rpart$best_metric, fwd_rf$best_metric,
                  fwd_svm$best_metric),
  bwd_metric  = c(bwd_lm$best_metric, bwd_tree$best_metric,
                  bwd_rpart$best_metric, bwd_rf$best_metric,
                  bwd_svm$best_metric),
  full_metric = c(
    AIC(lm(full_mtcars, data = mtcars)),
    deviance(tree(full_mtcars, data = mtcars)),
    rpart_xerror(rpart(full_mtcars, data = mtcars)),
    rf_oob_mse(randomForest(full_mtcars, data = mtcars)),
    svm_mse(svm(full_mtcars, data = mtcars))
  )
)
```

Forward and backward often select different subsets.  Forward starts sparse
and may get stuck once a mediocre predictor is in the model; backward starts
full and may keep predictors that only help marginally.  Neither dominates —
comparing both is a useful sanity check.

## Selected vs full formula

A natural question: does the search actually improve on just using every
predictor?  We can check by fitting each model family twice — once with the
`mine()`-selected formula and once with all predictors — and comparing on
the same metric.

To keep the candidate pool identical across methods, this comparison uses
first-order terms only (`max_degree = 1`, `max_interact_vars = 0`):

```{r vs-full-run, results = "hide"}
set.seed(42)
sel_lm <- mine(mtcars, mpg, max_degree = 1, max_interact_vars = 0)
sel_tr <- mine(mtcars, mpg, model_func = tree,   metric = deviance,
               max_degree = 1, max_interact_vars = 0)
sel_rp <- mine(mtcars, mpg, model_func = rpart,  metric = rpart_xerror,
               max_degree = 1, max_interact_vars = 0)
sel_rf <- mine(mtcars, mpg, model_func = randomForest, metric = rf_oob_mse,
               max_degree = 1, max_interact_vars = 0)
sel_sv <- mine(mtcars, mpg, model_func = svm, metric = svm_mse,
               max_degree = 1, max_interact_vars = 0)
```

```{r vs-full-compare}
full <- mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb

set.seed(42)
data.frame(
  model = c("lm (AIC)", "tree (deviance)",
            "rpart (xerror)", "randomForest (OOB MSE)",
            "svm (MSE)"),
  selected = c(
    AIC(lm(formula(sel_lm), data = mtcars)),
    deviance(tree(formula(sel_tr), data = mtcars)),
    rpart_xerror(rpart(formula(sel_rp), data = mtcars)),
    rf_oob_mse(randomForest(formula(sel_rf), data = mtcars)),
    svm_mse(svm(formula(sel_sv), data = mtcars))
  ),
  full = c(
    AIC(lm(full, data = mtcars)),
    deviance(tree(full, data = mtcars)),
    rpart_xerror(rpart(full, data = mtcars)),
    rf_oob_mse(randomForest(full, data = mtcars)),
    svm_mse(svm(full, data = mtcars))
  )
)
```

On mtcars the selected formula wins on every metric, including the
out-of-sample ones (`rpart` xerror is cross-validated; `randomForest` OOB
MSE is computed on held-out bags).  This is a small dataset (n = 32) with
heavy collinearity — exactly the setting where dropping redundant predictors
helps most.  On larger data with more independent signals the gap would
likely shrink.

## Synthetic data with a known interaction

To stress-test the comparison, we generate data where `y = x1 * x2 + noise`
with 6 noise predictors.  The true signal is a pure interaction, so model
functions that can represent interactions (explicitly or through splits)
have an advantage.

```{r synthetic-data}
set.seed(99)
n <- 200
x1 <- rnorm(n); x2 <- rnorm(n)
x3 <- rnorm(n); x4 <- rnorm(n)
x5 <- rnorm(n); x6 <- rnorm(n)
x7 <- rnorm(n); x8 <- rnorm(n)
dat <- data.frame(
  y = x1 * x2 + rnorm(n, sd = 0.5),
  x1, x2, x3, x4, x5, x6, x7, x8
)
```

```{r synthetic-run, results = "hide"}
set.seed(42)
syn_lm <- mine(dat, y, max_degree = 1, max_interact_vars = 2)
syn_tr <- mine(dat, y, model_func = tree, metric = deviance,
               max_degree = 1, max_interact_vars = 0)
syn_rp <- mine(dat, y, model_func = rpart, metric = rpart_xerror,
               max_degree = 1, max_interact_vars = 0)
syn_rf <- mine(dat, y, model_func = randomForest, metric = rf_oob_mse,
               max_degree = 1, max_interact_vars = 2)
syn_sv <- mine(dat, y, model_func = svm, metric = svm_mse,
               max_degree = 1, max_interact_vars = 2)
```

```{r synthetic-formulas}
data.frame(
  model   = c("lm", "tree", "rpart", "randomForest", "svm"),
  formula = c(deparse1(formula(syn_lm)), deparse1(formula(syn_tr)),
              deparse1(formula(syn_rp)), deparse1(formula(syn_rf)),
              deparse1(formula(syn_sv)))
)
```

Selected vs full on the same synthetic data:

```{r synthetic-vs-full}
syn_full <- y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8

set.seed(42)
data.frame(
  model = c("lm (AIC)", "tree (deviance)",
            "rpart (xerror)", "randomForest (OOB MSE)",
            "svm (MSE)"),
  selected = c(
    AIC(lm(formula(syn_lm), data = dat)),
    deviance(tree(formula(syn_tr), data = dat)),
    rpart_xerror(rpart(formula(syn_rp), data = dat)),
    rf_oob_mse(randomForest(formula(syn_rf), data = dat)),
    svm_mse(svm(formula(syn_sv), data = dat))
  ),
  full = c(
    AIC(lm(syn_full, data = dat)),
    deviance(tree(syn_full, data = dat)),
    rpart_xerror(rpart(syn_full, data = dat)),
    rf_oob_mse(randomForest(syn_full, data = dat)),
    svm_mse(svm(syn_full, data = dat))
  )
)
```

The results are more mixed than mtcars.  `lm` benefits dramatically because
it cannot represent the `x1 * x2` interaction unless the search discovers
it.  `randomForest` improves modestly with the explicit interaction term.
`tree` ties — it finds the same splits regardless of extra columns.
`rpart` does worse: the greedy search started from a noise predictor
(via the single-predictor fallback) and never recovered, while the full
formula gave `rpart` access to `x1` and `x2` from the start.

This illustrates a real limitation of greedy forward selection — the
starting point matters, and the fallback heuristic is not guaranteed to
pick a good one.

### Backward on synthetic data

Backward elimination avoids the starting-point problem because it begins
from all predictors.  On this dataset, `rpart` benefits — it starts with
`x1` and `x2` in the formula rather than depending on the single-predictor
fallback:

```{r synthetic-backward, results = "hide"}
set.seed(42)
syn_bwd_lm <- mine(dat, y, method = "backward",
                   max_degree = 1, max_interact_vars = 0)
syn_bwd_tr <- mine(dat, y, model_func = tree, metric = deviance,
                   method = "backward",
                   max_degree = 1, max_interact_vars = 0)
syn_bwd_rp <- mine(dat, y, model_func = rpart, metric = rpart_xerror,
                   method = "backward",
                   max_degree = 1, max_interact_vars = 0)
syn_bwd_rf <- mine(dat, y, model_func = randomForest, metric = rf_oob_mse,
                   method = "backward",
                   max_degree = 1, max_interact_vars = 0)
syn_bwd_sv <- mine(dat, y, model_func = svm, metric = svm_mse,
                   method = "backward",
                   max_degree = 1, max_interact_vars = 0)
```

```{r synthetic-backward-compare}
set.seed(42)
data.frame(
  model   = c("lm (AIC)", "tree (deviance)",
              "rpart (xerror)", "randomForest (OOB MSE)",
              "svm (MSE)"),
  forward  = c(syn_lm$best_metric, syn_tr$best_metric,
               syn_rp$best_metric, syn_rf$best_metric,
               syn_sv$best_metric),
  backward = c(syn_bwd_lm$best_metric, syn_bwd_tr$best_metric,
               syn_bwd_rp$best_metric, syn_bwd_rf$best_metric,
               syn_bwd_sv$best_metric),
  full     = c(
    AIC(lm(syn_full, data = dat)),
    deviance(tree(syn_full, data = dat)),
    rpart_xerror(rpart(syn_full, data = dat)),
    rf_oob_mse(randomForest(syn_full, data = dat)),
    svm_mse(svm(syn_full, data = dat))
  )
)
```

Backward elimination gives `rpart` a fair shot at the true signal and
typically matches or beats forward selection across model families on this
dataset.

## Notes

- **Automatic fallback.** `rpart()`, `randomForest()`, and `svm()` cannot
  fit intercept-only models.  `mine()` detects the failure and selects the
  best single-predictor starting point automatically.  Pass
  `keep_all_vars = TRUE` to start from all predictors instead, or use
  `method = "backward"` which always starts from all predictors.
- **Backward elimination.** `method = "backward"` starts from the full model
  and prunes down.  This avoids the intercept-only fallback entirely, making
  it especially useful for model functions that cannot fit sparse formulas.
- **No interactions for single trees.** `tree()` and `rpart()` reject
  interaction terms in the formula.  Use `max_interact_vars = 0` when
  working with these model functions.
- **OOB MSE varies across runs.** Random forests are stochastic; set a seed
  for reproducibility, or average over multiple seeds for more stable
  selection.
- **Deviance for `tree`.** The `deviance()` generic returns the residual
  sum of squares for regression trees.  It works as a built-in metric
  without a custom wrapper.
- **Cross-validation for `rpart`.** The `cptable` in an `rpart` model
  contains cross-validated error estimates.  The `rpart_xerror` wrapper
  above extracts the minimum, which accounts for tree complexity
  automatically.
- **SVM residual MSE.** `svm()` stores training residuals on the fitted
  object.  The `svm_mse` wrapper computes the mean squared residual.  This
  is a training metric — for honest evaluation, consider cross-validated
  error instead.
