---
title: "Using modelminer with randomForest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using modelminer with randomForest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
pkgload::load_all()
library(randomForest)
```

## Why feature selection for random forests?

Random forests handle high-dimensional data and interactions natively, so feature
selection is less critical than for linear models. Still, there are practical
reasons to search over term sets:

- **Interpretability.** A 3-predictor forest is easier to explain than a
  10-predictor forest.
- **Engineered features.** Explicit polynomial or interaction columns can
  sometimes help the tree splits, especially with small data.
- **Comparison.** Running the same `mine()` search on `lm` and `randomForest`
  shows how the two model families differ on the same data.

## Setup

`randomForest` accepts `formula` + `data`, so it plugs directly into `mine()`
as `model_func`. The one constraint is that `randomForest()` requires at least
one predictor — it cannot fit an intercept-only model — so we use
`keep_all_vars = TRUE` to start from all first-order terms.

For the metric, we use the OOB (out-of-bag) mean squared error that
`randomForest` computes automatically — no cross-validation loop needed:

```{r metric}
rf_oob_mse <- function(model) model$mse[model$ntree]
```

## Example: mtcars

```{r mtcars-run, results = "hide"}
set.seed(42)
rf_result <- mine(
  mtcars, mpg,
  model_func        = randomForest,
  metric            = rf_oob_mse,
  keep_all_vars     = TRUE,
  max_degree        = 2,
  max_interact_vars = 2
)
```

```{r mtcars-result}
rf_result$Formula
rf_result$best_metric
nrow(rf_result$all_models)
```

## Comparison: lm vs randomForest

We can run the same search with `lm` (using AIC) and compare which terms each
model family selects:

```{r compare, results = "hide"}
set.seed(42)
lm_result <- mine(
  mtcars, mpg,
  keep_all_vars     = TRUE,
  max_degree        = 2,
  max_interact_vars = 2
)
```

```{r compare-result}
data.frame(
  model   = c("lm (AIC)", "randomForest (OOB MSE)"),
  formula = c(deparse1(lm_result$Formula), deparse1(rf_result$Formula)),
  metric  = c(lm_result$best_metric, rf_result$best_metric)
)
```

## Synthetic data with a known interaction

To test whether the search finds a known signal, we generate data where
`y = x1 * x2 + noise` with 6 noise predictors:

```{r synthetic-data}
set.seed(99)
n <- 200
x1 <- rnorm(n); x2 <- rnorm(n)
x3 <- rnorm(n); x4 <- rnorm(n)
x5 <- rnorm(n); x6 <- rnorm(n)
x7 <- rnorm(n); x8 <- rnorm(n)
dat <- data.frame(
  y = x1 * x2 + rnorm(n, sd = 0.5),
  x1, x2, x3, x4, x5, x6, x7, x8
)
```

```{r synthetic-run, results = "hide"}
set.seed(42)
rf_syn <- mine(
  dat, y,
  model_func        = randomForest,
  metric            = rf_oob_mse,
  keep_all_vars     = TRUE,
  max_degree        = 1,
  max_interact_vars = 2
)
```

```{r synthetic-result}
rf_syn$Formula
rf_syn$best_metric
```

## Notes

- **`keep_all_vars = TRUE` is required.** `randomForest()` errors on an
  intercept-only formula, so the search must start with at least one predictor.
- **OOB MSE varies across runs.** The forest is stochastic; set a seed for
  reproducibility, or average over multiple seeds for more stable selection.
- **Interactions are implicit.** Random forests capture interactions through
  tree splits, so adding explicit `x1:x2` columns is redundant in many cases.
  The search is more useful for identifying which *variables* matter than for
  discovering interaction structure.
